---
title: "Distribution & CI"
author: "Yaochieh Yao"
date: "2/5/2023"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(mvtnorm, quietly = TRUE) #load library “mvtnorm”
data(golub, package = "multtest",quietly = TRUE)
```

## Module 4 Conte Carlo Simulation
**Problem 1 <br>**
Suppose that for certain microRNA of size 20 the probability of a purine is
binomially distributed with probability 0.7. Say there are 100 such microRNAs,
each independent of the other.<br>
<br>
Let Y denote the average number of purine in these microRNAs. Find the
probability that Y is great than 15. Please give a theoretical calculation, do NOT
use Monte Carlo simulation to approximate. Show all the steps and formulas in
your calculation.<br>
<br>
```{r}
# Since each sample is independent, CLT allows us to see them as a norm.
# P(Y>15) =1-P(Y<=15)
1-pnorm(15, mean=20*0.7,sd=sqrt(20*0.7*0.3)/sqrt(100)) #[1] 5.317746e-07
```
<br>
<br>
**Problem 2 <br>**
Two genes’ expression values follow a bivariate normal distribution. Let X and Y
denote their expression values respectively. Also, assume that X has mean=7 and
variance=3.Y has mean=12 and variance=7. The covariance between X and Y is 3.<br>
<br>
In a trial, 100 independent measurements of the expression values of the two genes
are collected, and denoted as , …,(X100, Y100). We wish to find the
probability , i.e., the probability that the sample mean for the
second gene exceeds the sample mean of the first gene more than 0.5.<br>
<br>
Conduct a Monte Carlo simulation to approximate this probability, providing a
95% confidence interval for your estimation. Submit your R script for the Monte
Carlo simulation.<br>
```{r}
# problem 2
nsim=1000
XmeanLess.sim<- rep(NA,nsim)
for (i in 1:nsim){
  data.sim<-rmvnorm(100,mean=c(7,12), sigma=matrix(c(3,3,3,7), nrow = 2))
  mean.sim<-apply(data.sim,2,mean)
  Xmean<-mean.sim[1]
  Ymean<-mean.sim[2]
  XmeanLess.sim[i]<-(Xmean+0.5<Ymean) 
}
mean(XmeanLess.sim) # [1] 1
mean(XmeanLess.sim)+c(-1,1)*1.96*sqrt(var(XmeanLess.sim)/nsim) #[1] 1 1
```
<br>
<br>
**Problem 3 <br>**
Assume there are three independent random variables X1~chisq(df~8),
X2~ Gamma(α=1, β=2), X3~ t-distribution with degrees of freedom m=5.<br>
Define a new random variable Y√X1X2+4(X3^2) as (note that the square
root is only for X1.)<br>
<br>
Use Monte Carlo simulation to find the mean of Y. Submit your R script for the
Monte Carlo simulation.
```{r}
x1<-rchisq(10000,df=8)
x2<-rgamma(10000,shape=1,scale=2)
x3<-rt(10000,df=5)
y<-sqrt(x1)*x2+4*(x3^2)
mean(y) #[1] 12.8268
```
<br>
<br>

**Problem 4 <br>**
Complete exercise 10 in Chapter 3 of Applied Statistics for Bioinformatics using R. Submit the plot, and a brief explanation of your observation.
The problem refers to the density function of extreme value distribution in another book. You do not have to look for the other book, the density function is <br>
<br>
        f(x)=exp(-x)exp(-exp(-x))<br>
<br>
Here exp(-x) is the same as e-x.<br>
<br>
ANS:<br>
The observed extreme value distribution approaches the f function pdf and
normal distribution for both the central location and wave scale.
```{r}
my.stat<-NULL
n<-10000
for (i in 1:1000) {  
  mx<-max(rnorm(n)) 
  an<-sqrt(2*log(n))-0.5*(log(log(n))+log(4*pi))*(2*log(n))^(-1/2)
  bn<-(2*log(n))^(-1/2)
  my.stat<-c(my.stat,(mx-an)/bn) 
}
plot(density(my.stat),frame=FALSE,main="Density plot",xlim=c(-4,5),ylim=c(0,0.7))
f<-function(x){exp(-x)*exp(-exp(-x))}
curve(f,range(density(my.stat)$x),add=TRUE,col = "blue")
curve(dnorm,add=TRUE,col = "red")
legend(x = "topleft", lty = c(2,3),bty = "n",
       col= c("black","blue","red"), 
       legend=c("Extreme value", "Function curve", "Normal distribution"))
```
<br>
<br>
<br>
<br>

## Module 5 Distribution & CI
**Problem 1 <br>**
A random sample of size 6 from the exp(λ) distribution results in observations:
1.433, 0.524, 0.384, 4.515, 1.852, 0.429. Find the MLE on this data set in two
ways:<br>
a.By numerical optimization of the likelihood (please include R code) and <br>
```{r warning=FALSE}
obs<-c(1.433,0.524,0.384,4.515,1.852,0.429)
# problem 1-a
lik<-function(lam) prod(dexp(obs, rate = lam))
nlik<-function(lam)-lik(lam)
optim(par=1, nlik)$par #[1] 0.6566895
```
<br>
<br>
b.by the analytic formula.<br>
```{r}
# problem 1-b
length(obs)/sum(obs)
```
<br>
<br>

**Problem 2 <br>**
A random sample X1, X2, ………., X75 follows chi-square distribution with m degree of
freedom, has sample mean 𝑋 " = 98.6 and sample standard deviation 𝑠 = 9.4.<br>
a.Find the point estimator of m using the method of moments.<br>
ANS:98.6<br>
<br>
<br>
b.Find a one-sided 90% lower confidence interval of m.<br>
Please provide the formulas and the derivations together with your numerical
answer.
```{r}
# problem 2-b 
n=75
98.6+qt(0.1,df=n-1)*9.4/sqrt(n) # (97.19645, Inf)
```
<br>
<br>

**Problem 3 <br>**
On the Golub et al. (1999) data set, analyze the Zyxin gene expression data
separately for the ALL and AML groups.<br>
a.Find the bootstrap 95% CIs for the mean and for the variance of the gene
expression in each group separately.
```{r}
gol.fac <- factor(golub.cl, levels=0:1, labels = c("ALL", "AML"))
AL_Zin<-golub[2124,gol.fac=="ALL"]
AM_Zin<-golub[2124,gol.fac=="AML"]
nboot<- 1000
AL_n<-length(AL_Zin)
AM_n<-length(AM_Zin)
# problem 3-a
AL.mean <- AL.var <- AM.mean <- AM.var <-rep(NA, nboot)
for (i in 1:nboot) { 
  AL.mean[i]<-mean(AL_Zin[sample(1:AL_n,replace=TRUE)]) 
  AM.mean[i]<-mean(AM_Zin[sample(1:AM_n,replace=TRUE)])
  AL.var[i]<-var(AL_Zin[sample(1:AL_n,replace=TRUE)])
  AM.var[i]<-var(AM_Zin[sample(1:AM_n,replace=TRUE)])
} 
quantile(AL.mean,c(0.025,0.975)) # -0.57641829 -0.04043985
quantile(AM.mean,c(0.025,0.975)) # 1.380224 1.803251
quantile(AL.var,c(0.025,0.975))  # 0.3551999 0.6506627
quantile(AM.var,c(0.025,0.975))  # 0.04554198 0.20368202 
```
<br>
<br>
b.Find the parametric 95% CIs for the mean and for the variance of the gene
expression in each group separately. (You need to choose the appropriate
approximate formula to use: z-interval, t-interval or chi-square interval.)
```{r}
# problem 3-b c(-1.96,1.96)
ci.all.mean<- mean(AL_Zin)+qt(c(0.025,0.975),df=AL_n-1)*sd(AL_Zin)/sqrt(AL_n) 
ci.aml.mean<-mean(AM_Zin)+qt(c(0.025,0.975),df=AM_n-1)*sd(AM_Zin)/sqrt(AM_n)
ci.all.var<-(AL_n-1)*sd(AL_Zin)^2/qchisq(c(0.975,0.025),df=AL_n-1)
ci.aml.var<-(AM_n-1)*sd(AM_Zin)^2/qchisq(c(0.975,0.025),df=AM_n-1)
print(ci.all.mean) #[1] -0.580738750 -0.008846435
print(ci.aml.mean) #[1] 1.339698 1.833638
print(ci.all.var) #[1] 0.3240441 0.9812951
print(ci.aml.var) #[1] 0.06597815  0.41621602
```
<br>
<br>
c.Find the bootstrap 95% CI for the median gene expression in both groups
separately.
```{r}
# problem 3-c
AL.median <- AM.median<- rep(NA, nboot)
for (i in 1:nboot){
  AL.median[i]<-median(AL_Zin[sample(1:AL_n,replace=TRUE)])
  AM.median[i]<-median(AM_Zin[sample(1:AM_n,replace=TRUE)])
}
quantile(AL.median,c(0.025,0.975)) # -0.73507  0.31432
quantile(AM.median,c(0.025,0.975)) # 1.22814 1.82829 
```
<br>
<br>
**Problem 4 <br>**
For a random sample of 50 observations from Poisson distribution, we have two
ways to construct a 90% CI for the parameter λ.<br>
1.Since the Poisson mean is λ, we can use the interval for the sample mean
(X ̄+t_0.05,49 √(X ̄/50),X ̄+t_0.95,49 √(X ̄/50))<br>
<br>
2.Since the Poisson variance is also λ, we can use the interval for the sample
variance directly: ((49s^2)/(χ_0.95,49^2 ),(49s^2)/(χ_0.05,49^2 )) <br>
<br>
a.Write a R-script to conduct a Monte Carlo study for the coverage probabilities
of the two CIs. That is, to generate nsim=1000 such data sets from the Poisson
distribution. Check the proportion of the CIs that contains the true parameter λ.<br>
```{r}
# problem 4 -a
MCsim<-function(nsim, lambda) {
  cov1<-cov2<-rep(NA,nsim)
  for (i in 1:nsim) {
    n=50
    x<- rpois(n,lambda)
    xbar<-mean(x)
    Xsd<-sd(x)
    CI1<-xbar+qt(c(.05,.95),df=n-1)*sqrt(xbar/n)
    CI2<-49*Xsd^2/qchisq(c(.95,.05), df=n-1)
    cov1[i]<-(CI1[1]<lambda)&(lambda<CI1[2])
    cov2[i]<-(CI2[1]<lambda)&(lambda<CI2[2])
  }
  print(paste("When lambda=", lambda, ": coverage for first CI is", mean(cov1), ", coverage for second CI is", mean(cov2), "."))
}
```
<br>
<br>
b.Run the Monte Carlo simulation for nsim=1000 runs, at three different
parameter values: λ=0.1, λ=1 and λ=10. Report the coverage probabilities of these
two CIs at each of the three parameter values.
```{r}
# problem 4 -b
MCsim(nsim=1000, lambda=0.1)  # (0.882,0.566)
MCsim(nsim=1000, lambda=1.0)  # (0.896,0.818)
MCsim(nsim=1000, lambda=10.0) # (0.902,0.899)
```

